Non-Vertical Cultural Transmission, Assortment, \\and the Evolution of Cooperation

### 2020-10-18 ms/main.tex
revise modifier model section

\author[1]{Dor Cohen}
\author[2]{Ohad Lewin-Epstein}
\author[3]{Marcus W. Feldman}
\author[1,4,5,*]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[4]{School of Zoology, Tel Aviv University, Tel Aviv, Israel}
\affil[5]{Sagol School of Neuroscience, Tel Aviv University, Tel Aviv, Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can reduce an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, including primates~\citep{jaeggi2013natural},  rats~\citep{rice1962altruism}, birds~\citep{stacey1990cooperative,krams2008experimental}, and lizards~\citep{sinervo2006self}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology~\citep[Appendix]{Haldane1932book}.

@@

Here, we study cultural-evolution models of cooperation that include both vertical and non-vertical transmission. 
We investigate these models using mathematical analysis and simulations.  
In our models behavioral changes are mediated by cultural transmission that can occur specifically during social interactions.
For instance, there may be an association between the choice of partner for social interaction and the choice of partner for cultural transmission.
As another example, when an individual interacts with an individual of a different phenotype,  exposure to the latter may lead the former to  convert its phenotype.
Our results demonstrate that cultural transmission can enhance the evolution of cooperation even when genetic transmission cannot, partly because it facilitates the generation of assortment \citep{Fletcher2009assortment}, and partly because non-vertical transmission can diminish the effect of natural selection \citep{ram2018evolution}.
This further emphasizes that treatment of cooperation as a cultural trait, rather than a genetic one, can lead to a broader understanding of its evolutionary dynamics.

@@

@@

If $p<p^*=1$, we want a condition for $1-p' < 1-p$, or
$\frac{1-p'}{1-p}<1$, or
$\lambda \frac{-(p'-p)}{1-p} < 0$, or
$-\frac{f(p)}{1-p} < 0$.
Using a linear approximation for $f(p)$ near $1$, we have
\begin{equation}\begin{aligned}
1-p' &< 1-p  \Leftrightarrow \\
\frac{f'(1)(p-1) + O\big((p-1)^2\big)}{p-1} &< 0 \Leftrightarrow \\
f'(1) - O(1-p) &< 0.
\end{aligned}\end{equation}
Therefore, if $f'(1)<0$ then there exists $\epsilon>0$ such that for any $1-\epsilon<1-p<1$ we have $1-p'<1-p$; that is, $p'$ is closer to one than $p$.
\end{appendices}

>>>>

\author[1]{Dor Cohen}
\author[2]{Ohad Lewin-Epstein}
\author[3]{Marcus W. Feldman}
\author[1,4,*]{Yoav Ram}

\affil[1]{School of Computer Science, Interdisciplinary Center Herzliya, Herzliya, Israel}
\affil[2]{School of Plant Sciences and Food Security, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Biology, Stanford University, Stanford, CA}
\affil[4]{School of Zoology, Tel Aviv University, Tel Aviv, Israel}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% Introduction
\section*{Introduction}
Cooperative behavior can reduce an individual's fitness and increase the fitness of its conspecifics or competitors~\citep{axelrod1981evolution}.
Nevertheless, cooperative behavior appears to occur in many non-human animals~\citep{dugatkin1997cooperation}, including primates~\citep{jaeggi2013natural},  rats~\citep{rice1962altruism}, birds~\citep{stacey1990cooperative,krams2008experimental}, and lizards~\citep{sinervo2006self}.
Evolution of cooperative behavior remains an important conundrum in evolutionary biology.

@@

Here, we study cultural-evolution models of cooperation that include both vertical and non-vertical transmission. 
We investigate these models using mathematical analysis and simulations.  
In our models behavioral changes are mediated by cultural transmission that can occur specifically during social interactions.
For instance, there may be an association between the choice of partner for social interaction and the choice of partner for cultural transmission.
As another example, when an individual interacts with an individual of a different phenotype,  exposure to the latter may lead the former to  convert its phenotype.
Our results demonstrate that cultural transmission can enhance the evolution of cooperation even when genetic transmission cannot, partly because it facilitates the generation of assortment \citep{Fletcher2009assortment}, and partly because non-vertical transmission can protect traits from the effect of natural selection \citep{ram2018evolution}.
This further emphasizes that treatment of cooperation as a cultural trait, rather than a genetic one, can lead to a broader understanding of its evolutionary dynamics.

@@

\subsection*{Modifier model}

We now analyze the evolution of social association under perfect vertical transmission, $v=1$.
We assume a population at a stable coexistence of the two phenotypes, cooperation $A$ and defection $B$, such that the frequency of $A$ among parents is $\tilde{p}^*$ (
\autoref{eq:vert_hori_equilibrium}).
Note that for stable coexistence to occur, we assume horizontal bias for defection, $T_A < T_B$, and an intermediate cost of cooperation, $\gamma_2 < c < \gamma_1$ (\autoref{eq:cost_boundaries})

The stable population mean fitness is $\bar{w}^* = 1 + \tilde{p}^*(b-c)$ (\autoref{eq:mean_fitness}), which is increasing in $\tilde{p}^*$. 
Therefore, to determine the effect of increasing $\alpha$ on the stable population mean fitness, we must analyze its effect on $\tilde{p}^*$, 
\begin{equation} \label{eq:mean_fitness_stable_coexistence_derivative}
  \frac{\partial \tilde{p}^*}{\partial \alpha} 
%  =\frac{-b^2T_A (1-\alpha) (T_A-T_B) + b(T_A-T_B)(c(1-T_B) - b \alpha T_A - (T_A-T_B))}{b^2 (1-\alpha)^2 (T_A-T_B)^2} 
%  = \frac{b^2T_A(T_B-T_A)-bc(1-T_B)(T_B-T_A) - b(T_B-T_A)^2}{b^2 (1-\alpha)^2 (T_A-T_B)^2} \;.
  = \frac{b T_A - c(1-T_B) + (T_A-T_B)}{b (1-\alpha)^2 (T_B-T_A)} \;.
\end{equation} 
Note that stable coexistence implies $c<\gamma_1$, and because $\alpha<1$, we have
\begin{equation}
c < \gamma_1 = \frac{b \alpha T_A + (T_A-T_B)}{1-T_B} < \frac{b T_A + (T_A-T_B)}{1-T_B}.
\end{equation} 
Therefore, $b T_A - c(1-T_B) - (T_B-T_A)>0$, and since $T_A<T_B$, the derivative $\partial \tilde{p}^* / \partial \alpha$ is positive, and the population mean fitness increases as social association $\alpha$ increases.

Therefore, if a modifier allele that increases social association invades the population, the population mean fitness will increase. But can such a modifier allele invade the population?
We assume that social association is determined by a genetically transmitted locus with alleles $m$ and $M$, which induce association probabilities $\alpha_1$ and $\alpha_2$, respectively.

We define the frequencies of the pheno-genotypes $Am$, $Bm$, $AM$, and $BM$ as $\hat{p}_1$, $\hat{p}_2$, $\hat{p}_3$, and $\hat{p}_4$, respectively. 
Then the following recursion system defines the frequencies of the pheno-genotypes in the next generation,
\begin{equation} \label{eq:next_gen_p_1}
  \begin{aligned}
	% p1
  \bar{w}\tilde{p}_1' & = \tilde{p}_1(\tilde{p}_1+\hat{p_3})(1+b-c)(1 - (1-\alpha_1)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_1T_B(\hat{p_1}+\tilde{p}_3) - T_B(\tilde{p}_2+\tilde{p}_4)) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_1(\tilde{p}_2 + \hat{p_4})) \\
  &+ \tilde{p}_2(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)T_A 
\\
	% p2
  \bar{w}\tilde{p}_2' & = \tilde{p}_2(\tilde{p}_2+\hat{p_4})(1 - (1-\alpha_1)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_2(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_1T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_1(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_1T_B + (1-\alpha_1)(\tilde{p}_2+\tilde{p}_4)T_B) \\
  &+ \tilde{p}_1(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_1)(\tilde{p}_2+\tilde{p}_4)T_B 
\\
	% p3
  \bar{w}\tilde{p}_3' & = \tilde{p}_3(\tilde{p}_1+\hat{p_3})(1+b-c)(1 - (1-\alpha_2)(\tilde{p}_2 + \tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(1-\alpha_2T_B(\hat{p_1}+\tilde{p}_3) - T_B(\tilde{p}_2+\tilde{p}_4)) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)T_A(\tilde{p}_1 + \tilde{p}_3 + \alpha_2(\tilde{p}_2 + \hat{p_4})) \\
  &+ \tilde{p}_4(\tilde{p}_2 + \tilde{p}_4)(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)T_A 
\\
	% p4
  \bar{w}\tilde{p}_4' & = \tilde{p}_4(\tilde{p}_2+\hat{p_4})(1 - (1-\alpha_2)(\tilde{p}_1 + \tilde{p}_3)T_A) \\
  &+ \tilde{p}_4(\tilde{p}_1 + \tilde{p}_3)(1+b)(1-\alpha_2T_A(\tilde{p}_2 + \tilde{p}_4)-T_A(\tilde{p}_1 + \tilde{p}_3))\\
  &+ \tilde{p}_3(\tilde{p}_2 + \tilde{p}_4)(1-c)(\alpha_2T_B + (1-\alpha_2)(\tilde{p}_2+\tilde{p}_4)T_B) \\
  &+ \tilde{p}_3(\tilde{p}_1 + \tilde{p}_3)(1-\alpha_2)(\tilde{p}_2+\tilde{p}_4)T_B \;,
  \end{aligned}
\end{equation}
where $\bar{w} = 1 + (b-c)\cdot(\tilde{p}_1 + \tilde{p}_3)$ is the population mean fitness.

Denote this system by $\bar{w} \vec{\tilde{p}} = f(\vec{\tilde{p}})$, where $\vec{\tilde{p}} = (\tilde{p}_1,\tilde{p}_2,\tilde{p}_3,\tilde{p}_4)$.
We are interested in the invasion of allele $M$ to a population at equilibrium with allele $m$, that is, with equilibrium frequencies $\vec{\tilde{p}}^*=(\tilde{p}^*, 1-\tilde{p}^*, 0, 0)$, with $\tilde{p}^*$ defined in
\autoref{eq:vert_hori_equilibrium}.
The success of the invasion of allele $M$ is determined by the external stability matrix $\mathbf{L}^*_{ex}$~\citep{Liberman1988}, which is derived from \autoref{eq:next_gen_p_1} as
\begin{equation}
\mathbf{L}^*_{ex} = \frac{1}{\bar{w}^*} \begin{bmatrix}
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_3'}{\partial \tilde{p}_4}(\vec{\tilde{p}}^*) \\
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_3}(\vec{\tilde{p}}^*) &
	 \frac{\partial \tilde{p}_4'}{\partial \tilde{p}_4} (\vec{\tilde{p}}^*)
\end{bmatrix} \;.
\end{equation}
If the leading eigenvalue of $\mathbf{L}^*_{ex}$ is less than 1, then the equilibrium $\vec{\tilde{p}}$ is locally stable and allele $M$ cannot invade the population of allele $m$.

@@

If $p<p^*=1$, we want a condition for $1-p' < 1-p$, or
$\frac{1-p'}{1-p}<1$, or
$\lambda \frac{-(p'-p)}{1-p} < 0$, or
$-\frac{f(p)}{1-p} < 0$.
Using a linear approximation for $f(p)$ near $1$, we have
\begin{equation}\begin{aligned}
1-p' &< 1-p  \Leftrightarrow \\
\frac{f'(1)(p-1) + O\big((p-1)^2\big)}{p-1} &< 0 \Leftrightarrow \\
f'(1) - O(1-p) &< 0.
\end{aligned}\end{equation}
Therefore, if $f'(1)<0$ then there exists $\epsilon>0$ such that for any $1-\epsilon<1-p<1$ we have $1-p'<1-p$; that is, $p'$ is closer to one than $p$.

\end{appendices}
